|

### 2021-08-12 dissertation/msc_dissertation.Rmd
Further writing

# Introduction
     The Covid-19 pandemic has reaffirmed the need for rigours scientific research to inform decision-making on a scientific and societal level. However, particularly the social sciences have a history of inflated positive findings leading to the reproducibility crisis [@pashler2012]. @simmons2011 demonstrated how the so-called *researcher degree of freedom* (RDF) affect study results. This concept refers to the analytical choices researchers make and their induced variation in results. For instance, sample size can be highly deterministic of finding statistically significant effects. @john2012 found 70% of researchers have at least once stopped their data collection based on a preliminary analysis. P-value simulations as a function of sample size showed an increased likelihood of obtaining false-positives for small sample sizes. Specifically, a sample size of ten had a 22.1% chance of being a false positive. Adding 20 observations per condition decreased the probability to 12.7% [@simmons2011]. The authors did not insinuate any malicious intent per-se, but rather attributed the issue to an ambiguity of a "right way" to analyse data, and needing statistically significant effects to publish. This example shows how easy it is to engineer these statistically significant effects. To prevent such bad practices @simmons2011 suggested requirements and guidelines for authors and reviewers.

 Among these requirements and guidelines are a minimum sample size, reporting all collected variables and presenting results with and without covariates [@simmons2011]. These are important suggestions but they struggle to cope with the full spectrum of the analytical possibilities. @silberzahn2018 further scrutinised the effects of RDF. Specifically, they had 29 teams of researchers investigate the effects of skin colour on the odds of being sent off the football field based on the same dataset. Other than being part of the project there was no incentive for researchers to participate i.e., there was no ulterior motive for researcher to manipulate outcomes (e.g. wanting/needing to publish). The variation between the analytic approaches was substantial. There were 29 different analyses with 21 different combinations of covariates. Twenty teams found a statistically significant effect; the odds ratios ranged from 0.89 to 2.93 (median = 1.31). The authors also controlled for researchers' prior believes and experience as well as peer rated analysis quality, none of them accounted for the variation of results. Other crowdsourced analysis projects have made similar observations [@botvinik-nezer2020]. These studies are interesting for two reasons (a) from a statistical perspective they highlight the RDF effects, and (b) from a content perspective they provide the most robust insights into the scrutinised concepts. Nonetheless, crowdsourcing approaches are impractical as they require extensive personnel (N number of researchers) and time resources (years) and are therefore often not feasible analytical strategies [@silberzahn2018].

 @silberzahn2018 and @botvinik-nezer2020 suggest using multiverse analysis, or specification-curve analysis, as an alternative to crowdsourcing approaches. This approach requires a researcher, ideally a more than one, to conduct with all plausible analyses, aggregate the results and conduct an inferential test across all possibilities to come to a conclusion. "Plausible" refers to statistically valid, non-redundant, tests appropriate to the research question [@simonsohn2020]. This approach is statistically more complex and computationally more intense than conventional analyses, but makes inherent strategy bias and noise transparent, and forces researchers to look at the entire "garden of forking paths" [@gelman2013]. A good example of quantifying the extend of variation due to model specification was conducted by @patel2015. They specified three modelling approaches (gaussian, cox and binomial) and 13 covariates. Their algorithm ran every possible covariate combination for each modelling approach which allowed the authors to assess the "vibration of effect" (VoE). The VoE is another term for quantifying the extend of variation in the results. The measure allows to assess the effect's robustness. The smaller the VoE, the less the effect depends on the analytical choices and vice versa. The authors therefore called for more caution when interpreting observational associations when finding large VoE's.

 Taken together, statistically significant effects are easy to engineer through adjusting analysis parameters like sample size. Crowdsourcing analysis presents a suitable strategy to understand the RDF induced variation but is impractical due to its substantial personnel and time requirements. Alternatives like multiverse analysis present a promising, low-personnel, but computationally intense alternative to identify RDF induced variation. The present research project will build upon the crowdsourcing findings of @silberzahn2018. Specifically, I will take the median analysis and rerun it with all possible covariate combinations. The objective is to define and understand the results space of the analysis i.e., assess the robustness of the median effect (VoE). Based on these insights I aim at contributing to further understanding the value, limitations and mechanisms of the multiverse approach.

# Analysis
## Analysis plan
     The project's objective is to define the results space of the @silberzahn2018 in order to draw inferences about the relationship between crowdsourcing and multiverse strategies. The results space refers to the numerical interval between the lowest and highest possible outcome. It is created by running every possible analytical strategy. @patel2015 refer to a related concept as the "Vibration of Effect" (VoE) and have developed a standardised approach to identifying this space. Essentially there are two factors: the analytical approach (i.e., the statistical model) and the covariates (or control variables). For @silberzahn2018 this means running every combination of 29 different analytical approaches and 15 different covariates used across all teams. These are $29 * 2^{15} = 950,272$ possibilities, which exceeds the scope of this research project. To get a more manageable number of possibilities of I will, therefore, focus on one analytical approach. In particular, on the approach that produced the median outcome of all analyses. The median outcome, being the middle number of any set of values, is a reasonable starting point in order to assess the VoE. Focusing on one analytical approach still leaves $1 * 2^{15} = 32,768$ possibilities. Due to computational limitations of my computer I will focus on a random sample of 1,000 thereof.

 In @silberzahn2018 the median outcome was produced by @team23 which will hereinafter be referred to as *team 23* due to it's designated team number in the original study. *Team 23* first transformed the data and then conducted a mixed-model logistic regression. To recreate the identical starting point for the analyses I will, first, apply the same transformations using the scripts provided by *team 23*. (The transformation process will be described in more detain the "data" section.) Second, I will create a matrix of all possible covariate combinations whereof a random sample of 1,000 is drawn. Third, because the original study investigated the relationship between football players' skin colour and the odds of being sent off the field, there will be a set of core variables included in all models to be able answer the original research question. Following on *team 23*, I will define two interaction terms as the model's core: "skin tone *X* implicit bias" and "skin tone *X* explicit bias". (The data set's variables will be described in more detail in the "data" section.) Forth, I will combine the core variables and the randomly drawn covariates as formulas to run the models. Fifth, after running the models I will visualise the outcomes as the conventional specification curve plot as well as a rain-cloud plot [@allen2021] and scatter plot. The aim here is develop easy to understand graphs that succinctly give insight into the results. Finally, I will discuss the implications for multiverse analysis.

>>>>

# Introduction
     The Covid-19 pandemic has reaffirmed the need for sound scientific research to inform decision-making on a scientific and societal level [@collins2021]. Rigorous research builds upon a systematic and well-reasoned approach to solving a research problem. Based on available literature, a falsifiable research question is defined and a corresponding hypothesis is developed. An appropriate study is then designed and conducted. To draw inferences from the gathered data, statistical models are developed and the effect of each variable is assessed. Every one of these steps is influenced by the decisions researchers make, which are known as *researcher degrees of freedom* [RDF, @simmons2011; @wicherts2016]. Usually there are many feasible analytical strategies to answering a research question and none of them are inherently right or wrong  [@carp2012]. This often creates uncertainty, for example, about what covariates to include and how to model them, which in turn leads to inconsistent findings [@patel2015; @ioannidis2008]. Recently, efforts have been made to better understand the scope of variation induced by different analytical strategies.

 These efforts include a crowdsourcing approach to data analysis [@silberzahn2018; @botvinik-nezer2020]. Its premise is to have a large number of researchers team up into to smaller, independent groups to investigate the same research question based on the same dataset. For instance, @silberzahn2018 had 29 teams investigate the effects of a football player's skin colour on the odds of being sent off the field. The variation between the analytical strategies was substantial. There were 29 different analyses with 21 different combinations of covariates. Twenty teams found a statistically significant effect. The authors also controlled for researchers' prior believes and experience as well as peer-rated analysis quality, none of which accounted for the variation of results. @botvinik-nezer2020 made similar observations. Crowdsourcing data analysis excels at emphasising the substantial impact of different analytical strategies, but has a significant drawback. It is extremely time and personnel intensive, the two mentioned studies lasted between 2 to 3+ years, and included 61 and 180 analysts, respectively. Not to mention the organisational effort.

 @silberzahn2018 and @botvinik-nezer2020, therefore, suggest using multiverse analysis (also known as specification-curve analysis) as an alternative to crowdsourcing. This approach requires identifying and running all plausible analytical strategies. Plausible strategies are defined as statistically valid, non-redundant, tests appropriate to the research question. Their aggregated results are then used to make inferences about the research question. Despite being statistically more complex and computationally more intense, this approach has the advantages of only needing one, or ideally a couple, researchers. Moreover, a researcher's inherent strategy bias is neutralised and noise is made transparent [@simonsohn2020]. Multiverse approaches are therefore well suited for assessing the scope of variation induced by different analytical strategies. For instance, @patel2015 assessed an effect's robustness by running every combination of covariates and modelling options. Their conclusion was the larger the variation of outcomes, the less robust is the effect, and the less it should therefore to be trusted. The authors termed this measure the *vibration of effect* (VoE).

 Taken together, researchers make many analytical decisions throughout conduction a study which are know as *researcher degrees of freedom*. Crowdsourcing analysis presents a strategy to understand their induced variation, but is impractical due to its substantial time and personnel requirements. Multiverse analysis present a promising, low-personnel, but computationally intense alternative to assessing the scope this variation. Despite observing substantial variation in crowdsourced data analysis projects little is known about the extend to which these approaches cover the full range of possible results. In this project, using methods of multiverse analysis I will define the results space of @silberzahn2018 to estimate study's representativeness. Based on these insights I aim at contributing to further understanding the value, limitations and mechanisms of the multiverse approach.

# Analysis
## Analysis plan
     The project's objective is to define the results space of the @silberzahn2018 to the estimate study's representativeness. The results space refers to the numerical interval between the lowest and highest possible outcome. It is created by running every possible analytical strategy. Hence, it is closely related to assessing an effect's robustness. @patel2015 developed a standardised approach to assessing an effect's robustness which they termed assessing the *vibration of effect*. I will therefore use this standardised approach to define the results space. The approach essentially comprises two parameters: the statistical models and the covariates (or control variables). In @silberzahn2018 the analysts used numerous different statistical models like multiple linear regression, mixed-model logistic regression or Bayesian logistic regression. In total there were 29 different modelling approaches. Additionally, each team used a different set of covariates. Across all teams there 15 covariates used. This gives $2^n$ i.e., $2^{15} = 32,768$ possible combinations of covariates. Adding all modelling possibilities to the equation gives a total of $29 * 2^{15} = 950,272$ combinations to run. This number exceeds the project's scope, hence, it needs to be reduce to a more manageable count. I am, therefore, focusing on one modelling approach. In particular, on the approach that produced the median outcome of all analyses. The median, being the middle number of any set of values, is a reasonable starting point in order to estimate the results space. Nevertheless, even focusing on one analytical approach still leaves $1 * 2^{15} = 32,768$ possibilities. Due to computational limitations I will first run a sample of 200, followed by a sample of 1,000 combinations. If there's a substantial difference between the median and the spread I will run more models, if not, I will make the assumption that 1,000 reasonably estimates all ~33k combinations.

 In @silberzahn2018 the median outcome was produced by @team23 which will hereinafter be referred to as *team 23* due to it's designated team number in the original study. *Team 23* first transformed the data and then conducted a mixed-model logistic regression. The first step of this project will be replicating the transformation and the analysis. Replicating other researchers analyses has the benefit of checking their work and, if results are indeed replicated, increasing confidence in them. It also ensures that this project has the same starting point as *team 23* did. Given the team made all their scripts publicly available I expect this step to be straightforward. Next is drawing a random sample of covariates, without replacement. "Without replacement" ensures all covariate combinations are unique in the sample. The covariates will be appended to the base (or core) variables. Base variables are those variables that are primarily assessed to answer the research question. In this case whether a football player's skin colour effects the odds of being sent off the field. *Team 23* defined two interaction terms as the base: "skin tone *X* implicit bias" and "skin tone *X* explicit bias". (The variables are described in the "data" section.) Hence, all models will have the following structure:
$$
\begin{aligned}
RedCard = SkinTone*ImplicitBias + SkinTone*ExplicitBias + CovariateCombination_{i}
\end{aligned}
$$ 
While running each model, the relevant statistics will be extracted. Those are the coefficient (or effect) of skin tone, its standard error, test-statistic and p-value. Just as *team 23* did I will then calculate the 95% confidence intervals (CI). The estimates and their CIs are then transformed to odds ratios (OR) through exponentiating them to the power of two. Odds ratios quantify the strength of association between two variables. If greater than one the dependent variable is more likely to occur given the independent variable, if lower than one it less likely. The odds ratios will be visualised to assessed through a conventional specification curve plot, a rain-cloud plot and scatter plot. The following describes them in more detail:

(i) The specification curve plot was developed to be a descriptive plot i.e., raw data is being plotted without any aggregation done. Its objective is describe the results space while allowing the reader to identify the specified covariate for each outcome [@simonsohn2020]. It therefore has two vertically stacked components. The top part shows a sorted scatter plot. On the horizontal axis are the models specifications and on the vertical axis is the outcome measure. The points are sorted from lowest to highest outcome measure. This way the lowest outcome measure is on the bottom left corner and the highest in the top right one. The bottom part of the plot represents are table. Just like the top the horizontal axis hold the model specifications, the vertical axis lists the covariates. This arrangement allows each cell to specify the presence or absence for each covariate in a given specified model. For the top and bottom plot to work together it is imperative that the horizontal axes are identically arranged. If so, for every point in the top plot (i.e., for every outcome) the specified covariates can be seen in the bottom plot. For an example check the results section. The issue, however, is that if too many specifications are on the horizontal axis it becomes very hard to identify specific covariates. Hence, I also included the rain-cloud and scatter plot.

(ii) The rain-cloud plot was developed by @allen2021. Its objective is to give an unbiased, transparent view on the raw data. It therefore combines a density, box and scatter plot. If stacked vertically from top to bottom, respectively, they look like a cloud with rain drops, hence, its name. The advantage of this plot is to be able to assess the raw data including key statistics (median, 25th and 75th quartiles and CIs) as well as the probability density all in one succinct plot. Given that this project seeks to define and describe a results space the rain-cloud plot the ideal graph to do this. However, the plot neither allows to identify the covariates giving rise to the outcome nor their specified effects. (For an example check the results section.) Therefore a third plot is developed which is to be assessed in tandem with this plot.

(iii) The third and final plot is a sorted scatter plot. While there is nothing special about the plot itself its content is important. Given the very high number of models that are run, it does not make sense to assess each model and its effects in isolation. Instead the effects for each covariates neeed to be aggregated. Hence, the goal of this plot is to viusalise the specified effect for each covariate. Prior to doing so, the coviates first need to be aggregated and their specified effects need to be calucualted. To this end, each covariate is recoded into a binary factor: included in the model yes/no. This newly defined factors are than used as an independent variables in an ANOVA. The previously calcualted odds rations are used as dependent variables. The estimates of the fitted model are the specified effect for each covariates. These are visualised similarly to the top part of the specification curve plot. The covariates are on the horizental axis, will the odds ratios are on the vertical axis. Here too, the lowest oucome is in the bottom left corner and the highest outcome in the top right one. Additionally the point are coloured indicating a statistically significant effect (or non-signifcant).
